environmental_data_preperation

### 2018-09-19 src/environmental_data_preperation.Rmd
NAO data done

# Overview
Preperation of GLORYS and BIOMER netCDFs obtained from [Copernicus](www.marine.copernicus.eu/services-portfolio/access-to-products/). These databases give global oceanographic conditions.

@@

# read in a netcdf
note this is a GLORYS dataset
```{r}
library("ncdf4")
ncbr <- brick("../data/env/CMEMS_GLOBAL_REANALYSIS_PHY_001_025_salinity_monthly_mean_1993-01.nclonlat.ncdepthtrim.nc")
ncbr
```

# reproject 
There is currently an issue with the projectRaster command in the raster package. Michael Sumners kindly offered a [temporary solution](https://twitter.com/mdsumner/status/1040500830539935744)
```{r}
## use ncdf4 directly
library(ncdf4)
## connection to the file
nc <- nc_open("../data/env/CMEMS_GLOBAL_REANALYSIS_PHY_001_025_salinity_monthly_mean_1993-01.nclonlat.ncdepthtrim.nc")
## desired variable 
v <- ncvar_get(nc, "salinity")
## dimension coordinates
lon <- ncvar_get(nc, "longitude")
lat <- ncvar_get(nc, "latitude")
## a template, one map from the lon, lat, and v[,,1]
template <- raster(list(x = lon, y = lat, z = v[,,1]), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 ")

@@

#if projectRaster would work properly...

@@

#create unique id layer
ok now you want to create a duplicate of the top layer and then give each cell an unique ID

@@

-----------------------------------------
#The Biomer Layers...

@@

#read in a biomer netcdf
```{r}
library("ncdf4")
biobrick <- brick("../data/env/CMEMS_GLOBAL_REANALYSIS_BIO_001_018_chl_monthly_mean_1998-01.ncwgs.nc_lonlat.ncdepthtrim.nc")
biobrick
```

#Project the biomer file to match the glorys layer
```{r}
bio_aea <- projectRaster(biobrick, aea) #note this only projects the top layer
bio_aea
writeRaster(bio_aea,'../output/env/bio_aea.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
```

@@

#AMO data 
Downloaded from [NOAA](https://www.esrl.noaa.gov/psd/data/timeseries/AMO/) (AMO, unsmoothed, short) as a txt and then converted to a .csv

@@

And just for fun, plot the AMO... follow the tutorial [here](http://www.stat.pitt.edu/stoffer/tsa4/R_toot.htm)
```{r}
amocolumn <- matrix(t(amoraw[ ,2:13]), ncol=1) #month data is in col 2 - 13
amocolumn1 <- as.vector(amocolumn)
amotimeseries <- ts(amocolumn1, frequency = 12, start = c(1998,1))
plot.ts(amotimeseries)
abline(h = 0)
dev.copy(png, "../output/env/amo.png") #this prints a png of the plot
dev.off() #this turns off the print command
```

@@

>>>>

# Overview
Preperation of GLORYS and BIOMER netCDFs obtained from [Copernicus](www.marine.copernicus.eu/services-portfolio/access-to-products/). These databases give global oceanographic condition.
UPDATE: also includes prep of the AMO and NAO data...

@@

# NetCDFs (Biomer and GLORYS)
read in a netcdf
note this is a GLORYS dataset
```{r}
library("ncdf4")
ncbr <- brick("../data/env/CMEMS_GLOBAL_REANALYSIS_PHY_001_025_salinity_monthly_mean_1993-01.nclonlat.ncdepthtrim.nc")
ncbr
```

reproject ...
There is currently an issue with the projectRaster command in the raster package. Michael Sumners kindly offered a [temporary solution](https://twitter.com/mdsumner/status/1040500830539935744)
```{r}
## use ncdf4 directly
library(ncdf4)
## connection to the file
nc <- nc_open("../data/env/CMEMS_GLOBAL_REANALYSIS_PHY_001_025_salinity_monthly_mean_1993-01.nclonlat.ncdepthtrim.nc")
## desired variable 
v <- ncvar_get(nc, "salinity")
## dimension coordinates
lon <- ncvar_get(nc, "longitude")
lat <- ncvar_get(nc, "latitude")
## a template, one map from the lon, lat, and v[,,1]
template <- raster(list(x = lon, y = lat, z = v[,,1]), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 ")

@@

if projectRaster would work properly...

@@

# create unique id layer
ok now you want to create a duplicate of the top layer and then give each cell an unique ID

@@

-----------------------------------------
# The Biomer Layers...

@@

read in a biomer netcdf
```{r}
library("ncdf4")
biobrick <- brick("../data/env/CMEMS_GLOBAL_REANALYSIS_BIO_001_018_chl_monthly_mean_1998-01.ncwgs.nc_lonlat.ncdepthtrim.nc")
biobrick
```

Project the biomer file to match the glorys layer
```{r}
bio_aea <- projectRaster(biobrick, aea) #note this only projects the top layer
bio_aea
writeRaster(bio_aea,'../output/env/bio_aea.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
```

@@

# AMO data

Downloaded from [NOAA](https://www.esrl.noaa.gov/psd/data/timeseries/AMO/) (AMO, unsmoothed, short) as a txt and then converted to a .csv

@@

And just for fun, plot the AMO... follow the tutorial [here](http://www.stat.pitt.edu/stoffer/tsa4/R_toot.htm)
```{r}
amocolumn <- matrix(t(amoraw[ ,2:13]), ncol=1) #month data is in col 2 - 13
amocolumn1 <- as.vector(amocolumn)
amotimeseries <- ts(amocolumn1, frequency = 12, start = c(1996,1))
plot.ts(amotimeseries)
abline(h = 0)
dev.copy(png, "../output/env/amo.png") #this prints a png of the plot
dev.off() #this turns off the print command
```

@@

# NAO

Dataset mes from [NCAR](https://climatedataguide.ucar.edu/climate-data/hurrell-north-atlantic-oscillation-nao-index-pc-based). Note I am using the monthly principal component (PC)-based indices of the North Atlantic Oscillation [(NAO)](https://climatedataguide.ucar.edu/sites/default/files/nao_pc_monthly.txt). I have already converted this file into a csv and removed unncesessary text. Note there are no headers...

```{r}
naoraw <- read.csv("../data/env/amon.us.data.csv", header = FALSE) 
head(naoraw)
```

lets give some column names. col 1  year, 2:13 are the months
```{r}
colnames(naoraw) <- c("year","1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12")
head(naoraw)
```

 NA values are represented by -99.99. Change these to NA
```{r}
naoraw <- replace(naoraw,naoraw==-99.99, NA)
```

Now subset the dataset to start in 1996 (the capelin data starts in 1998, but you want some previous time period stuff)

```{r}
nao <- naoraw[naoraw$year >= 1996, ]
head(nao) 
```

And just for fun, plot the NAO follow the tutorial here

```{r}
naocolumn <- matrix(t(amoraw[ ,2:13]), ncol=1) #month data is in col 2 - 13
naocolumn1 <- as.vector(naocolumn)
naotimeseries <- ts(naocolumn1, frequency = 12, start = c(1996,1))
plot.ts(naotimeseries)
abline(h = 0)
dev.copy(png, "../output/env/nao.png") #this prints a png of the plot
dev.off() #this turns off the print command
```

Spring:	1 March - 31 May (col 4:6) Summer: 1 June - 31 August (col 7:9) Autumn: 1 Sept - 31 Nov (col 10:12) Winter: 1 Dec - 28 Feb (col 13 from prev year, 1:2)

```{r}
naophases <- nao 
naophases$YearAvg <- rowMeans(naophases[2:13], na.rm = TRUE) #the month data is in col 2-13
naophases$SpringAvg <- rowMeans(naophases[4:6], na.rm = TRUE)
naophases$SummerAvg <- rowMeans(naophases[7:9], na.rm = TRUE)
naophases$AutumnAvg <- rowMeans(naophases[10:12], na.rm = TRUE)
```

winter has a different format - the columns are in 1 & 2, & 13 from the previous column, which means the rowmeans function needs a bit of tweaking...

shift the dec column you created down by 1
```{r}
library(useful)
naophases <- shift.column(data = naophases, columns = "12", len = 1, up = FALSE)
naophases
write.csv(naophases, file = "../output/env/nao.csv", row.names = FALSE) #for some reason you need to write to a csv and then reload for the next step to work properly...

```

he shifted data is in 12.Shifted...
```{r}
naophases <- read.csv("../output/env/nao.csv")
naophases$WinterAvg <- rowMeans(naophases[ , c("X12.Shifted", "X1", "X2")], na.rm = TRUE)  #x because apparently all the col names start with x for some reason...
naophases
```

Now label the phases positive, negative (or avg if any happen to be at 0)

```{r}
naophases$YearPhase[naophases$YearAvg ==0] <- "avg"
naophases$YearPhase[naophases$YearAvg >0] <- "positive"
naophases$YearPhase[naophases$YearAvg <0] <- "negative"
naophases$SpringPhase[naophases$SpringAvg ==0] <- "avg"
naophases$SpringPhase[naophases$SpringAvg >0] <- "positive"
naophases$SpringPhase[naophases$SpringAvg <0] <- "negative"
naophases$SummerPhase[naophases$SummerAvg ==0] <- "avg"
naophases$SummerPhase[naophases$SummerAvg >0] <- "positive"
naophases$SummerPhase[naophases$SummerAvg <0] <- "negative"
naophases$AutumnPhase[naophases$AutumnAvg ==0] <- "avg"
naophases$AutumnPhase[naophases$AutumnAvg >0] <- "positive"
naophases$AutumnPhase[naophases$AutumnAvg <0] <- "negative"
naophases$WinterPhase[naophases$WinterAvg ==0] <- "avg"
naophases$WinterPhase[naophases$WinterAvg >0] <- "positive"
naophases$WinterPhase[naophases$WinterAvg <0] <- "negative"
```

now save this as a new csv

```{r}
write.csv(naophases, file = "../output/env/nao.csv", row.names = FALSE)
```
